Efficient ancestry and mutation simulation with msprime 1.0

### 2021-09-14 paper.tex
Add Broad affiliation for Daniel Goldstein

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{listings}
\usepackage[english]{babel}%
\usepackage[T1]{fontenc}%
\usepackage[utf8]{inputenc}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{dsfont}
\usepackage{verbatim}%
\usepackage{environ}%
\usepackage[right]{lineno}%
\usepackage{nameref}
\usepackage{caption,subcaption}
\usepackage{tikz}
\usepackage[right]{lineno}
\usetikzlibrary{calc,positioning}
%\usepackage{showkeys}

@@

\newcommand{\jkcomment}[1]{\textcolor{red}{#1}}
\newcommand{\plrcomment}[1]{\textcolor{cyan}{#1}}
\newcommand{\grgcomment}[1]{\textcolor{yellow!60!red}{#1}}
\newcommand{\adkcomment}[1]{\textcolor{blue}{#1}}

@@

\linenumbers
\title{Efficient ancestry and mutation simulation with msprime 1.0}
% First authors
\author[1,$\star$]{Franz Baumdicker}
\author[2,$\star$]{Gertjan Bisschop}
\author[3,$\star$]{Daniel Goldstein}
\author[4,$\star$]{Graham Gower}
\author[5,$\star$]{Aaron P. Ragsdale}
\author[6,$\star$]{Georgia Tsambos}
\author[7,$\star$]{Sha Zhu}

% Middle Authors
\author[8]{Bjarki Eldon}
\author[9]{Castedo E. Ellerman}
\author[10,11]{Jared G. Galloway}
\author[12,13]{Ariella L. Gladstein}
\author[14]{Gregor Gorjanc}
\author[15]{Bing Guo}
\author[7]{Ben Jeffery}
\author[16]{Warren W. Kretzschmar}
\author[17]{Ivan Krukov}
\author[2]{Konrad Lohse}
\author[18]{Michael Matschiner}
\author[17]{Dominic Nelson}
\author[19]{Nathaniel S. Pope}
\author[20]{Consuelo D. Quinto-Cort\'es}
\author[10]{Murillo F. Rodrigues}
\author[21]{Kumar Saunack}
\author[22]{Thibaut Sellinger}
\author[23]{Kevin Thornton}
\author[9]{Hugo van Kemenade}
\author[7,24]{Anthony W. Wohns}
\author[7]{H. Yan Wong}

% Senior Authors
\author[16,$\dagger$]{Simon Gravel}
\author[10,$\dagger$]{Andrew D. Kern}
\author[25,$\dagger$]{Jere Koskela}
\author[10,26,$\dagger$]{Peter L. Ralph}

@@

% Franz
\affil[1]{Cluster of Excellence ``Controlling Microbes to Fight Infections'',
Mathematical and Computational Population Genetics, University of T\"ubingen}
% Gertjan, Konrad
\affil[2]{Institute of Evolutionary Biology, The University of Edinburgh}
% Daniel
\affil[3]{Khoury College of Computer Sciences, Northeastern University}
% Graham
\affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
% Aaron
\affil[5]{Department of Integrative Biology, University of Wisconsin--Madison}
% Georgia
\affil[6]{Melbourne Integrative Genomics, School of Mathematics and Statistics,
    University of Melbourne}
% Jerome, Yan, Wilder, Joe
\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
    University of Oxford}

% Bjarki
\affil[8]{Leibniz Institute for Evolution and Biodiversity
Science, Museum f\"ur Naturkunde Berlin }
% Castedo, Hugo
\affil[9]{No affiliation}
% Jared, Murillo, Peter, Andy
\affil[10]{Institute of Ecology and Evolution, Department of Biology,
    University of Oregon}
% Jared
\affil[11]{Computational Biology Program, Fred Hutchinson Cancer Research
Center, Seattle, WA 98102, USA}
% Ariella
\affil[12]{Department of Genetics, University of North Carolina at Chapel Hill}
\affil[13]{Embark Veterinary, Inc., Boston}
% Gregor
\affil[14]{The Roslin Institute and Royal (Dick) School of Veterinary Studies,
    University of Edinburgh}
% Bing
\affil[15]{Institute for Genome Sciences, University of Maryland School of
Medicine, Baltimore, MD}
% Winni
\affil[16]{Center for Hematology and Regenerative Medicine, Karolinska Institute}
% Simon, Dom, Ivan
\affil[17]{Department of Human Genetics, McGill University}
% Michael
\affil[18]{Natural History Museum, University of Oslo}
% Nathaniel
\affil[19]{Department of Entomology, Pennsylvania State University}
% Consuelo
\affil[20]{National Laboratory of Genomics for Biodiversity (LANGEBIO),
    Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
% Saunack
\affil[21]{IIT Bombay, India}
% Thibaut
\affil[22]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
% Kevin
\affil[23]{Ecology and Evolutionary Biology, University of California Irvine}
% Wilder 2
\affil[24]{Broad Institute of MIT and Harvard}
% Jere
\affil[25]{Department of Statistics, University of Warwick}
% Peter (second affil)
\affil[26]{Department of Mathematics, University of Oregon}

@@

\begin{abstract}
Stochastic simulation is a key tool in population genetics,
since the models involved are often analytically intractable
and simulation is usually the only way of obtaining ground-truth data to
evaluate inferences.
Because of this necessity, a large number
of specialised simulation programs have been developed, each
filling a particular niche, but with largely overlapping functionality
and a substantial duplication of effort.
Here, we introduce \msprime\ version 1.0, which
efficiently implements ancestry and mutation simulations based on
the succinct tree sequence data structure and \tskit\ library.
We summarise \msprime's many features, and show that
its performance is excellent, often many times faster
and more memory efficient than specialised alternatives.
These high-performance features have been thoroughly
tested and validated, and built using a collaborative,
open source development model,
which reduces duplication of effort and promotes
software quality via community engagement.
% "Bazaar" development model: https://en.wikipedia.org/wiki/The_Cathedral_and_the_Bazaar
\end{abstract}

\textbf{Keywords:} Simulation, Coalescent, Mutations, Ancestral Recombination
Graphs

@@

The coalescent
process~\citep{kingman1982coalescent,kingman1982genealogy,hudson1983testing,
tajima1983evolutionary}
models the ancestry of a set of sampled genomes,
providing a mathematical description of the
genealogical tree that relates the samples to one another.
It has proved to be a powerful model,
and is now central to
population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
The coalescent is an efficient framework for population genetic
simulation, because it allows us to simulate the genetic ancestry for
a sample from an idealised population model, without explicitly representing
the population in memory or stepping through the generations.
Indeed, \cite{hudson1983testing} independently derived the
% Anyone disagree here? That's the impression I get from the paper.
coalescent \emph{in order to} efficiently simulate data,
and used these simulations to characterise an analytically
intractable distribution.
This inherent efficiency, and the great utility of simulations for a wide
range of purposes, has led to dozens of different tools
being developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
peng2015genetic}.

Two technological developments of recent years, however,
pose major challenges to most existing simulation methods.
Firstly, fourth-generation sequencing technologies have made
complete chromosome-level assemblies possible~\citep{miga2020telomere},
% Some citations here? Darwin tree of life maybe?
and high quality assemblies are now available for many species.
Thus, modelling genetic variation data as a series of unlinked
non-recombining loci
is no longer a reasonable approximation, and
we must fully account for recombination.
However, while a genealogical tree relating $n$ samples in the single-locus coalescent
can be simulated in $O(n)$ time~\citep{hudson1990gene},
the coalescent with recombination is far more complex,
and programs such as
Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination.
The second challenge facing simulation methods is that
sample sizes in genetic studies have grown very quickly
in recent years, enabled by the precipitous fall in genome sequencing costs.
Human datasets like the
UK Biobank~\citep{bycroft2018genome} and
gnomAD~\citep{karczewski2020mutational} now consist of hundreds of
thousands of genomes and many other datasets on a similar scale
are becoming available~\citep{tanjo2021practical}.
Classical simulators such as \ms\ and even fast approximate methods
such as \scrm~\citep{staab2015scrm} simply cannot
cope with such a large number of samples.

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has greatly increased the scope of coalescent simulations,
and it is now straightforward to simulate millions of whole chromosomes
for a wide range of organisms.
The ``succinct tree sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
originally introduced as part of \msprime, makes it possible to store
such large simulations in a few gigabytes, several orders
of magnitude smaller than commonly used formats.
The succinct tree sequence has also led to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic
statistics~\citep{kelleher2016efficient,ralph2020efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making it a highly efficient and flexible platform for population
genetic simulation.
This paper marks the release of \msprime\ 1.0.
We provide an overview of its extensive features,
demonstrate its performance advantages over alternative software,
and discuss opportunities for ongoing
open-source community-based development.

% and finally lay out future plans for
% further open-source development of what we hope will become an industry
% standard piece of computational infrastructure for the genetics community.

\section*{Results}
In the following sections we describe the main features of \msprime\ 1.0,
focusing on the aspects that are either new for this version, or in which
our approach differs significantly from classical methods. Where appropriate,
we benchmark \msprime\ against other simulators, but the comparisons are
illustrative and not intended to be systematic or exhaustive. Please
see~\cite{kelleher2016efficient} for a performance comparison of
\msprime\ against simulators such as \ms, \msms, and \scrm.

% jk: I started writing a roadmap here, but it's quite hard to do concisely
% and you either end up leaving sections out, or just summarising them
% one-by-one which makes it long and tedious. I think readers can either read
% through fully if they're interested, or just skim the section titles which
% should give them a good idea of whether they want to read it or not.

@@

\begin{figure}
    \includegraphics{illustrations/mutated_tree}
    \caption{
        \label{fig-mutated-trees}
        Visualisation of the separation between ancestry and mutation
        simulation. (A) The result of an invocation of \texttt{sim\_ancestry}
        is two trees along a 1kb chunk of genome relating three diploid samples.
        Each diploid individual consists of two genomes (or nodes), indicated
        by colour.
        (B) This ancestry is provided as the input to \texttt{sim\_mutations},
        which adds mutations.
        Graphics produced using \tskit's \texttt{draw\_svg} method.
    }
\end{figure}

A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
designed in part to avoid sources of error (see the Demography section) but
also to provide more appropriate defaults while keeping compatibility with
existing code. In the new APIs, ancestry and mutation simulation are fully
separated (see Fig.~\ref{fig-mutated-trees}),
with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
functions replacing the legacy \texttt{simulate} function. Among other changes,
the new APIs default to discrete genome coordinates and finite sites mutations,
making the default settings more realistic and resolving a major source of confusion and error.
The previous APIs are fully
supported and tested, and will be maintained for the foreseeable future.
The \texttt{msp} program has been extended to include new commands
for simulating ancestry and mutations separately. A particularly useful
feature is the ability to specify demographic models in
Demes format~\citep{gower2021demes} from the command line,
making simulation of complex demographies straightforward.
We also provide an \ms\ compatible
command line interface to support existing workflows.

\begin{figure}
\begin{center}
\includegraphics{illustrations/example_tree_sequence}
\end{center}
\caption{\label{fig-ts-example} An example tree sequence describing
genealogies and sequence variation for four samples at ten sites
on a chromosome of twenty bases long.
Information is stored in a set of tables
(the tables shown here include only essential columns, and
much more information can be associated with the various entities).
The node table
stores information about sampled and ancestral genomes. The
edge table describes how these genomes are related along a
chromosome, and defines the genealogical tree at each position.
The site and mutation tables together describe sequence variation
% JK: also amongst the ancestral genomes, but only in positions where
% they inherit from the sample. I thought it would be less confusing
% to leave this out?
among the samples.
}
\end{figure}

One of the key reasons for \msprime's substantial performance advantage
over other simulators~\citep{kelleher2016efficient}
is its use of the ``succinct tree sequence''
data structure to represent simulation results.
The succinct tree sequence (usually abbreviated to ``tree sequence'')
was introduced by \cite{kelleher2016efficient}
to concisely encode genetic ancestry and sequence variation
and was originally implemented as part of \msprime.
We subsequently extracted the core tree sequence functionality
from \msprime\ to create the \tskit\ library,
which provides a large suite of tools for processing genetic ancestry
and variation data via APIs
in the Python and C languages~\citep{tskit2021tskit}.
The availability of \tskit\ as a liberally licensed (MIT) open source
toolkit has enabled several other
projects~\citep[e.g.][]{kelleher2019inferring,haller2019slim,wohns2021unified,
de2021geonomics} to take advantage
of the same efficient data structures used in \msprime,
and we hope that many more will follow.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we summarise
some aspects that are important for simulation.

>>>>

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{listings}
\usepackage[english]{babel}%
\usepackage[T1]{fontenc}%
\usepackage[utf8]{inputenc}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{dsfont}
\usepackage{verbatim}%
\usepackage{environ}%
\usepackage[right]{lineno}%
\usepackage{nameref}
\usepackage{tikz}
%\usepackage{showkeys}

@@

\newcommand{\jkcomment}[1]{\textcolor{red}{#1}}
\newcommand{\plrcomment}[1]{\textcolor{cyan}{#1}}
\newcommand{\grgcomment}[1]{\textcolor{yellow!60!red}{#1}}

@@

\title{Efficient ancestry and mutation simulation with msprime 1.0}
% First authors
%% FIXME affiliations are all broken
\author[1,$\star$]{Franz Baumdicker}
\author[2,$\star$]{Gertjan Bisschop}
\author[3,24,$\star$]{Daniel Goldstein}
\author[4,$\star$]{Graham Gower}
\author[5,$\star$]{Aaron P. Ragsdale}
\author[6,$\star$]{Georgia Tsambos}
\author[7,$\star$]{Sha Zhu}

% Middle Authors
\author[8]{Bjarki Eldon}
\author[9]{Castedo E. Ellerman}
\author[10]{Jared G. Galloway}
\author[11]{Ariella L. Gladstein}
\author[12]{Gregor Gorjanc}
\author[13]{Guo, Bing}
\author[7]{Ben Jeffery}
\author[14]{Warren W. Kretzschmar}
\author[15]{Ivan Krukov}
\author[2]{Konrad Lohse}
\author[16]{Michael Matschiner}
\author[15]{Dominic Nelson}
\author[17]{Nathaniel Pope}
\author[18]{Consuelo D. Quinto-Cort\'es}
\author[10]{Murillo F. Rodrigues}
\author[19]{Kumar Suanack}
\author[20]{Thibaut Sellinger}
\author[21]{Kevin Thornton}
\author[22]{Hugo van Kemenade}
\author[7]{Anthony W. Wohns}
\author[7]{H. Yan Wong}

% Senior Authors
\author[15,$\dagger$]{Simon Gravel}
\author[10,$\dagger$]{Andrew Kern}
\author[23,$\dagger$]{Jere Koskela}
\author[10,24,$\dagger$]{Peter L. Ralph}

@@

% Franz
\affil[1]{Mathematical and Computational Population Genetics, University of T\"ubingen}
% Gertjan, Konrad
\affil[2]{Institute of Evolutionary Biology, The University of Edinburgh}
% Daniel
\affil[3]{Khoury College of Computer Sciences, Northeastern University}
% Graham
\affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
% Simon, Dom, Ivan
\affil[5]{FIXME AARON's AFFILIATION}
% Georgia
\affil[6]{Melbourne Integrative Genomics, School of Mathematics and Statistics,
    University of Melbourne}
% Jerome, Yan, Wilder, Joe
\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
    University of Oxford}

% Bjarki
\affil[8]{Leibniz Institute for Evolution and Biodiversity
Science, Museum f\"ur Naturkunde Berlin }
% Castedo
\affil[9]{FIXME: CASTEDO's AFFILIATION}
% Jared, Murillo, Peter, Andy
\affil[10]{Department of Biology and Institute of Ecology and Evolution,
    University of Oregon}
% Ariella
\affil[11]{Department of Genetics, University of North Carolina at Chapel Hill}
% Gregor
\affil[12]{The Roslin Institute and Royal (Dick) School of Veterinary Studies,
    University of Edinburgh}
% Bing
\affil[13]{FIXME: BING's AFFILIATION}
% Winni
\affil[14]{FIXME: WINNI's AFFILIATION}
% Simon, Dom, Ivan
\affil[15]{Department of Human Genetics, McGill University}
% Michael
\affil[16]{FIXME: MICHAEL's AFFILIATION}
% Nathaniel
\affil[17]{FIXME: NATHANIEL's AFFILIATION}
% Consuelo
\affil[18]{National Laboratory of Genomics for Biodiversity (LANGEBIO),
    Unit of Advanced Genomics, CINVESTAV, Irapuato, Mexico}
% Saunack
\affil[19]{FIXME: SAUNACK's AFFILIATION}
% Thibaut
\affil[20]{Professorship for Population Genetics, Department of Life Science Systems, Technical University of Munich}
% Kevin
\affil[21]{Ecology and Evolutionary Biology, University of California at Irvine}
% Hugo
\affil[22]{FIXME HUGO's AFFILIATION}
% Jere
\affil[23]{Department of Statistics, University of Warwick}
% Peter (second affil)
\affil[24]{Department of Mathematics, University of Oregon}

@@

\begin{abstract}
Coalescent simulation is a key tool in population genetics and
has been integral to theoretical developments since its beginnings.
Because of the ease with which the basic coalescent model can be simulated,
a large number of different simulators have been developed. However,
the coalescent with recombination is far more challenging to simulate,
and, until recently, it was not possible to simulate efficiently.
The \msprime\ software has revolutionised
coalescent simulation, making it possible to simulate millions
of whole genomes for the first time. We summarise the many features
of \msprime\ 1.0, which is built around a core model of efficiently
implementing recombination via the succinct tree sequence data
structure. Moreover, development of \msprime\ is collaborative
and open source, which reduces duplication of effort and promotes
software quality via community engagement.
% "Bazaar" development model: https://en.wikipedia.org/wiki/The_Cathedral_and_the_Bazaar
\end{abstract}

\textbf{Keywords:} Coalescent simulation, Python, ancestral recombination graphs

@@

The coalescent
process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
models the ancestry of a set of sampled genomes. Specifically,
the coalescent provides a mathematical description of the
genealogical tree that relates sampled genomes to one another.
The coalescent has proved to be a powerful model,
and it is now central to
population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
Among the most impactful contributions of the coalescent since its origins
~\citep{hudson1983testing} is that it provides a path towards more efficient
simulation of population genetic samples than is possible using traditional,
forwards in time simulation.
As a result of this great utility, dozens of different simulation tools have been
developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
peng2015genetic}.
Two technological developments of recent years, however,
pose major challenges to the majority of these methods.
% Some citations here? Darwin tree of life maybe?
Firstly, whole genome sequencing---and in particular fourth-generation
sequencing technologies---have made chromosome-level assemblies
commonplace. Thus, modelling genetic variation data as a series of unlinked loci,
a previously reasonable assumption, is no longer defensible and
we must fully account for recombination.
While a genealogical tree relating $n$ samples in the single-locus coalescent
can be simulated in $O(n)$ time~\citep{hudson1990gene},
the coalescent with recombination, which generates multiple correlated trees, is
far more challenging.
% JK will need to revise this sentence, in light of what we'll be saying
% down further about needing the SMC for very large rho.
Programs such as Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination,
and, until recently, approximate models of
recombination~\citep{mcvean2005approximating,staab2015scrm}
were necessary to simulate whole chromosomes.

The second major challenge to existing coalescent simulators is the
exponential increase in sample size of genetic studies. Human datasets like the
UK Biobank~\citep{bycroft2018genome} and
gnomAD~\citep{karczewski2020mutational} now consist of hundreds of
thousands of genomes and many other datasets on a similar scale
are becoming available~\citep{tanjo2021practical}.
Classical simulators such as \ms\ and even fast approximate methods
such as \scrm\ simply cannot cope with such a large number of samples.

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has greatly increased the scope of coalescent simulations,
and it is now straightforward to simulate millions of whole chromosomes.
The ``succinct tree sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
introduced as part of \msprime\, makes it possible to store
such huge simulations in a few gigabytes, several orders
of magnitude smaller than commonly used formats.
The succinct tree sequence has also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic
statistics~\citep{kelleher2016efficient,ralph2020efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making it a highly efficient and flexible platform for population
genetic simulation.
This paper marks the release of \msprime\ 1.0, a state-of-the-art
coalescent simulator both with respect to efficiency as well as feature-richness.
Here we provide an overview of the extensive abilities of \msprime\ 1.0,
demonstrate its performance advantages over alternative software,
and finally lay out future plans for
further open-source development of what we hope will become an industry
standard piece of computational infrastructure for the genetics community.

\section*{Results}

\jkcomment{Give a quick roadmap,
like, we discuss the underlying tree sequence object,
then mutations (because it's quite separate from the rest)
and then a bunch of new features for ancestry simulation. There's no
point in listing it out section by section.}

% JK: commented out this text as it's misleading and would confuse
% people. We should adapt it when writing the roadmap when the structure
% has settled down.

% In this section we describe the main features of \msprime\ 1.0. Methodologically,
% there exists a logical separation between simulating coalescent genealogies and
% adding neutral mutations onto these genealogies. Simulating genetic ancestry
% requires methods that model the underlying reproductive process (e.g. Moran,
% Wright-Fisher, etc.) as well as accounting for recombination of various sorts
% (including gene conversion), demography, and selection. These methods are
% discussed in the first N subsections below. One of the strengths of the
% approach taken in msprime is that neutral mutations can be retrospectively
% added to simulated genealogies. Methods to generate mutations, and thereby
% generate genetic variation, are therefore discussed in a separate and subsequent
% subsection. Finally, we describe the \msprime\ API, which is designed to enforce
% the distinction between simulating genealogies and mutations; discuss the
% interoperability provided by the use of \tskit\, the tree sequence toolkit;
% and highlight the open source development process which allows for \msprime\
% to be extended by addition of new features while ensuring robustness and
% maintainability of the core codebase.

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{illustrations/example_tree_sequence}
\end{center}
\caption{\label{fig-ts-example} Example tree sequence.
An example tree sequence describing genealogies and genotypes
for four samples at ten sites.
}
\end{figure}

One of the key reasons for \msprime's substantial performance advantage
over other simulators~\citep{kelleher2016efficient}
is its usage of the ``succinct tree sequence''
 data structure to represent simulation results.
The succinct tree sequence (usually abbreviated to ``tree sequence'')
is a data structure for concisely encoding
genetic ancestry and sequence variation.
The basic concepts were introduced by \cite{kelleher2016efficient}
and originally developed as part of \msprime. We subsequently
extracted the tree sequence code that was independent of simulations
from \msprime\ into
a standalone Python and C library, called \tskit~\citep{tskit2021tskit}.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we summarise
some aspects that are important for simulation.

% Encoding ancestry
Given a set of sample chromosomes, the topology
of their genetic ancestry is described by a set of \emph{nodes}
and \emph{edges} (Fig.~\ref{fig-ts-example}).
A node in tree sequence terminology corresponds
to a single chromosome strand, associated with its birth time
and other optional information. A node may be
associated with an \emph{individual}, so that, for example, a diploid
individual is associated with two separate nodes (we omit
the individual table in Fig.~\ref{fig-ts-example} as it is optional).
Nodes define haplotypes that exist at a specific time within
some individual, and edges define genetic inheritance
between these nodes. An \emph{edge} consists of a
parent node, a child node and the left and right coordinates
of the contiguous genomic interval over which
the child inherited genetic material from the parent.
Critically, edges can span multiple marginal trees: for
example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral chromosomes and recording the
genomic intervals over which parent/child relationships exist,
generalises the original
``coalescence records'' concept~\citep{kelleher2016efficient},
and is the key to the efficiency of tree
sequences~\citep{
kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.

The tree sequence encoding of ancestry as a set of nodes and edges
as just described, is equivalent to a graph in which edges
are annotated with a set of disjoint genomic intervals.
This is closely related to the Ancestral Recombination Graph (ARG),
a term that is often loosely understood
to mean some representation of genetic
ancestry as a graph~\citep[e.g.][]{mathieson2020ancestry}.
In this sense a succinct tree sequence \emph{is} an ARG.
More technically, an ARG is a particular
representation of genetic ancestry as a graph based on the
flow of ancestral material through recombination and common
ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
In this sense, tree sequences can be used to \emph{encode}
ARGs without loss of information (although
the default minimal tree sequence encoding is
much more concise; see the \nameref{sec-arg} section).
It should be noted that, despite standardisation
efforts~\citep{cardona2008extended,mcgill2013graphml}, there is
no shared format for interchanging ARG data, and no supporting software
libraries.

% OK, back to tree sequences: what about sequence variation?
The final output of most population genetic simulations is some representation
of sequence variation among the specified samples. For coalescent
simulations, we usually have three steps: (1) simulate
the genetic ancestry, and optionally output the resulting marginal trees;
(2) simulate sequence evolution conditioned on this ancestry by generating
mutations (see the \nameref{sec-mutations} section); and (3) output the
resulting sequences by percolating the effects of the mutations through
the trees. Information about the mutations
themselves---e.g., where they have occurred on the
trees---is usually not retained or made available for subsequent analysis.
In \msprime, however, we skip step (3), instead using \tskit's
combined data model of ancestry and mutations
to represent the simulated sequences.
As illustrated in Fig.~\ref{fig-ts-example}, mutations are a
fully integrated part of \tskit's tree sequence data model, and genetic
variation is encoded by recording sites at which mutations
have occurred, and where each mutation at those sites
has occurred on the marginal tree.
Crucially, the genetic sequences themselves are never stored, or indeed directly
represented in memory (unless explicitly decoded using \tskit\ APIs).
It may at first seem inconvenient to have only this indirect representation
of the genome sequences, but it is extremely powerful.
Firstly, the storage space required for simulations is dramatically
reduced. For a simulation of $n$ samples with $m$ variant sites,
we would require $O(nm)$ space to store the sequence data as a variant
matrix. However, if this simulation was of a recombining genome
with $t$ trees, then the \tskit\ tree sequence encoding
requires $O(n + t + m)$ space,
assuming we have $O(1)$ mutations at each site~\citep{kelleher2016efficient}.
For large sample sizes, this difference is profound, making it
conceivable, for example, to store the genetic ancestry
and variation data for the entire human population on a
laptop~\citep{kelleher2019inferring}. As well as the huge
difference in storage efficiency, it is often far more efficient
to compute statistics of the sequence data from the
trees and mutations than it is to work with the sequences themselves.
For example, computing Tajima's $D$ from simulated data stored in the \tskit\ format
is several orders of magnitude faster than efficient variant matrix
libraries for large sample sizes~\citep{ralph2020efficiently}.

The software ecosystem built around \tskit\ is stable and mature;
simulators such as
\fwdpy~\citep{thornton2014cpp},
\SLiM~\citep{haller2019slim},
\stdpopsim~\citep{adrion2020community},
\texttt{Geonomics}~\citep{de2021geonomics}
and
\texttt{GSpace}~\citep{virgoulay2021gspace},
and inference methods such as
\texttt{tsinfer}~\citep{kelleher2019inferring},
\texttt{tsdate}~\citep{,wohns2021unified}
and \texttt{Relate}~\citep{speidel2019method}
use either the Python or C
APIs to support outputting results in tree sequence format.
Tree sequences are stored in an efficient binary file format, and
are fully portable across operating systems and processor
architectures. The \tskit\ library ensures interoperability
between programs by having strict definitions of how the
information in each of the tables is interpreted, and
stringent checks for the internal consistency of the data model.

@@

A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
designed in part to avoid sources of error (see the Demography section) but
also to provide more appropriate defaults while keeping compatibility with
existing code. In the new APIs, ancestry and mutation simulation are fully
separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
functions replacing the legacy \texttt{simulate} function. Among other changes,
the new APIs default to discrete genome coordinates and finite sites mutations,
making the default settings more realistic and resolving a major source of confusion and error.
The previous APIs are fully
supported and tested, and will be maintained for the foreseeable future.
We also provide an \ms\ compatible
command line interface to support existing workflows. The
\texttt{msp} program has been extended to include new commands
for simulating ancestry and mutations separately. A particularly useful
feature is the ability to specify demographic models in
Demes format~\citep{gower2021demes} from the command line,
making simulation of complex demographies straightforward.

% The output - what do we get out?
\subsection*{Data analysis}
\label{sec-data-analysis}
The point of running simulations is to generate data, and there
are two important ways in which the \tskit\ tree sequences output by
\msprime\ are more useful than classical methods:
they are more efficient and they contain more information.
We will consider these points in turn.

The standard way of representing simulation data is
to render the results in a text format. For example,
\ms~\citep{hudson2002generating} outputs a set of sequences
and can also optionally output the marginal trees along
the genome in Newick format,
and variants of this approach are used by many simulators.
Text files have many advantages, but are slow to process
at scale.
The ability to efficiently process simulation results is
particularly important in simulation-based inference methods
such as Approximate Bayesian Computation
(ABC)~\citep{beaumont2002approximate,csillery2010approximate,wegmann2010abctoolbox}
and machine learning based
approaches~\citep{sheehan2016deep,chan2018likelihood,schrider2018supervised,
flagel2019unreasonable,sanchez2020deep}. Clearly, simulation efficiency is
crucial since the size and number of simulations that can be performed determines
the depth to which one can sample from the model and parameter space.
Equally important,
however, is the efficiency with which the simulation results can be
transformed into the specific input required by the inference method.
In the case of ABC, this is usually a set of summary statistics of the sequence
data, and methods avoid the bottleneck of parsing
text-based file formats to compute these statistics
by either developing their own
simulators~\citep[e.g.][]{cornuet2008inferring,lopes2009popabc}
or creating forked versions (i.e., modified copies) of existing
simulators~\cite[e.g.][]{thornton2006approximate,
hickerson2007msbayes,pavlidis2010msabc,huang2011mtml,quinto2018modeling},
tightly integrated with the inference method.
Modern approaches to ABC such as
ABC-RF~\citep{raynal2019abc,pudlo2016abc} and
ABC-NN~\citep{csillery2012abc,blum2010abc} use large
numbers of weakly informative statistics,
making the need to efficiently compute statistics from simulation
results all the more acute.

Classical text based output formats like \ms\ are inefficient to process,
but also lack a great deal of important information.
The tree-by-tree topology information output by simulators in Newick
format lacks any concept of node identity,
and means that we cannot reliably infer information about ancestors
from the output. Because Newick stores branch lengths rather
than node times, numerical precision issues also arise when summing
over many short branches~\citep{mcgill2013graphml}.
Numerous forks  of simulators have been created to access information not provided
in the output. For example, \ms\  has been forked to
output information about migrating segments~\citep{rosenzweig2016powerful},
information about ancestral lineages~\citep{chen2013asymptotic},
an array of population genetic statistics~\citep{ramos2007mlcoalsim}
% Any more?
and \ms's fork \msHOT~\citep{hellenthal2007mshot},
has in turn been forked to output information on local
ancestry~\citep{racimo2017archaic}.
All of this information
is either directly available by default in \msprime, or can be optionally
stored via options such as \texttt{record\_migrations} or
\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
can be efficiently and conveniently processed via \tskit\ APIs.

These problems of inefficient and limited access to the simulated
data are comprehensively solved by \msprime\ and \tskit. Complete
information about ancestry and mutation in simulations is immediately
available to the user in a single integrated data structure
(rather than multiple independent files), with no intermediate
processing required.
Users can subsequently compute precisely the information that they
require using the powerful suite of population genetic statistics
and other methods available in \tskit.
